---
title: "Statistical Modeling"
output: html_document
date: "2025-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dlnm)
library(splines)
library(sp)
library(spdep)
library(sf)
```

## Load dataset

```{r}
geo_location <- read.csv("./lon_lat.csv")
colnames(geo_location)[1] <- "NAME_1"
Japan.shapefile <- st_read('./Japan_shp/JPN_adm1.shp')
Japan.shapefile <- sp::merge(Japan.shapefile,geo_location,by="NAME_1",duplicateGeoms = TRUE)
Japan.shapefile <- as(Japan.shapefile, "Spatial")
rsv_data <- read.csv("./full data.csv")
rsv_data <- rsv_data[,-1]
rsv_data$tscluster <- factor(rsv_data$tscluster)
rsv_data$NAME_1 <- factor(rsv_data$NAME_1)
```
## Construct adjacency matrix

```{r}
adj_list <- Japan.shapefile %>% st_as_sf() %>% st_touches()
names(adj_list) <- Japan.shapefile$NAME_1
```

## Meteorological Model

### Covariates

```{r}
# covariates
cb.vis <- crossbasis(rsv_data$visibility,lag=3,argvar=list(fun="ps",df=5),arglag=list(fun="ps",degree=5),group=rsv_data$NAME_1)  
cb.sp <- crossbasis(rsv_data$sea.level.pressure,lag=3,argvar=list(fun="ps",df=5),arglag=list(fun="ps",degree=5),group=rsv_data$NAME_1)
```

### Exposures

```{r}
cb.tavg.piecewise <- crossbasis(rsv_data$TAVG,lag=3,argvar=list(fun="bs",knots=c(10,20),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.rh.piecewise <- crossbasis(rsv_data$relative.humidity,lag=3,argvar=list(fun="bs",knots=c(60,70),degree=1),arglag=list(fun="poly",degree=3),group= rsv_data$NAME_1)
cb.ws.piecewise <- crossbasis(rsv_data$wind.speed,lag=3,argvar=list(fun="bs",knots=c(20),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)  
cb.trange <- crossbasis(rsv_data$TRange,lag=3,argvar=list(fun="lin"),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.p <- crossbasis(rsv_data$PRCP,lag=3,argvar=list(fun="poly",degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)  
cb.hd <- crossbasis(rsv_data$Days.Over.30,lag=3,argvar=list(fun="lin"),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.cd <- crossbasis(rsv_data$Days.Below.0,lag=3,argvar=list(fun="lin"),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
```

### Crude Association (use average temperature as an example to illustrate)

```{r}
crude_tavg_model <- gam(Cases~cb.tavg.piecewise+s(tscluster,bs="re")+ # ts-cluster
                          s(week,by=Year,k=30)+ # temporal trend
                          s(NAME_1,bs="mrf", k=40,xt=list(nb=adj_list)), # spatial variation
                     family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.tavg.25 <- crosspred(cb.tavg.piecewise,crude_tavg_model,cen=25,cumul=TRUE,by=1)
```

### Univariate Association (use average temperature as an example to illustrate)

```{r}
univariate_tavg_model <- gam(Cases~cb.tavg.piecewise+s(tscluster,bs="re")+s(week,by=Year,k=30)+s(NAME_1,bs="mrf", k=40,xt=list(nb=adj_list))+
                               s(cb.vis)+s(cb.sp), # covariates 
                     family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.tavg.25 <- crosspred(cb.tavg.piecewise,univariate_tavg_model,cen=25,cumul=TRUE,by=1)
```

### Multivariate Association (use average temperature as an example to illustrate)

```{r}
multivariate_tavg_model <- gam(Cases~cb.tavg.piecewise+cb.rh.piecewise+cb.ws.piecewise+cb.p+cb.trange+cb.hd+cb.cd+# exposures
                                 s(tscluster,bs="re")+s(week,by=Year,k=30)+s(NAME_1,bs="mrf",k=40,xt=list(nb=adj_list))+
                                 s(cb.vis)+s(cb.sp),
                     family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.tavg.25 <- crosspred(cb.tavg.piecewise,multivariate_tavg_model,cen=25,cumul=TRUE,by=1)
```

## Air Pollutant Model

### Covariates

```{r}
# covariates
cb.sp <- crossbasis(rsv_data$sea.level.pressure, lag=3, argvar=list(fun="ps",df=5), arglag=list(fun="ps",degree=5),
                     group= rsv_data$NAME_1)
cb.rh <- crossbasis(rsv_data$relative.humidity, lag=3, argvar=list(fun="ps",df=5), arglag=list(fun="ps",degree=5),
                     group= rsv_data$NAME_1)
cb.tavg <- crossbasis(rsv_data$TAVG, lag=3, argvar=list(fun="ps",df=5), arglag=list(fun="ps",degree=5),
                     group= rsv_data$NAME_1)
cb.ws <- crossbasis(rsv_data$wind.speed, lag=3, argvar=list(fun="ps",df=5), arglag=list(fun="ps",degree=5),
                     group= rsv_data$NAME_1)
```

### Exposures

```{r}
cb.pm25 <- crossbasis(rsv_data$PM25,lag=3,argvar=list(fun="bs",knots=c(15,50),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.no2 <- crossbasis(rsv_data$NO2,lag=3,argvar=list(fun="bs",knots=c(25,50,120),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.co <- crossbasis(rsv_data$CO,lag=3,argvar=list(fun="bs",knots=c(4),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)
cb.so2 <- crossbasis(rsv_data$SO2,lag=3,argvar=list(fun="bs",knots=c(40),degree=1),arglag=list(fun="poly",degree=3),group=rsv_data$NAME_1)  
```

### Crude Association (use PM2.5 as an example to illustrate)

```{r}
crude_pm25_model <- gam(Cases~cb.pm25+s(tscluster,bs="re")+ # ts-cluster
                          s(week,by=Year,k=30)+ # temporal trend
                          s(NAME_1,bs="mrf",k=40,xt=list(nb=adj_list)), # spatial variation
                        family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.pm25.50 <- crosspred(cb.pm25,crude_pm25_model,cen=50,cumul=TRUE,by=1)
```

### Univariate Association (use PM2.5 as an example to illustrate)

```{r}
univariate_pm25_model <- gam(Cases~cb.pm25+s(tscluster,bs="re")+s(week,by=Year,k=30)+s(NAME_1,bs="mrf",k=40,xt=list(nb=adj_list))+
                               s(cb.rh)+s(cb.sp)+s(cb.tavg)+s(cb.ws), # covariates
                             family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.pm25.50 <- crosspred(cb.pm25,univariate_pm25_model,cen=50,cumul=TRUE,by=1)
```

### Multivariate Association (use PM2.5 as an example to illustrate)

```{r}
multivariate_model <- gam(Cases~cb.pm25+cb.so2+cb.co+cb.no2+ # exposures
                                 s(tscluster,bs="re")+s(week,by=Year,k=30)+s(NAME_1,bs="mrf",k=40,xt=list(nb=adj_list))+
                                 s(cb.rh)+s(cb.sp)+s(cb.tavg)+s(cb.ws),
                             family=quasipoisson(),data=rsv_data,method='REML',select=TRUE)
pred.pm25.50 <- crosspred(cb.pm25,multivariate_model,cen=50,cumul=TRUE,by=1)
```

